A Meta-analysis of Longevity Estimates of Mosquito Vectors of Disease: Supplementary Material

### 2018-11-23 MS.tex
Sorted table references

Data from MRR experiments in the Guerra et al. (2014) database were
examined and those with fewer than six recaptures and species with only
a single MRR study were excluded for the hierarchical analysis. Of the
232 data sets, 179 involved only females, 35 males, and 18 both sex
releases. For 102 data sets the age of the released mosquitoes was known
(the average age of released mosquitoes was 4.0 days) while in the other
cases it was unknown or unrecorded; in these cases we assumed the
mosquitoes were newly emerged at the time of release and return to this
assumption later. See Table S\ref{S-tab:mrr_aggregateData} for a summary of other data characteristics.

@@

The estimated mean LBL for female mosquitoes not pre-fed with blood or
sugar before release for \emph{Culex, Anopheles} and \emph{Aedes} were
2.9, 6.8 and 8.1 days respectively with an overall estimate of 6.0 days
(Fig. \ref{fig:mrr_lifespans}; Table S\ref{S-tab:mrr_estimated_lifespans}). Differences between genera were significant (ANOVA on median
LBL controlling for sex and pre-release feeding:
\emph{F}\textsubscript{2,229} = 12.4, \emph{p}\textless{}0.01; Kruskal
Wallace: \({\chi^{2}}_{2} = 30.8,p < 0.01\)). \emph{K}-fold cross
validation suggests that after the effect of genus is accounted for the
incorporation of a species term provides little predictive power (Fig.
S\#; in part explained by the latter model over-fitting the data where
there are few time series per species).

@@

Dissection allows the number of completed gonotrophic cycles to be
counted and from this the mean number of cycles before death was
estimated. Overall, the mean number of cycles completed in a lifetime
was 1.3 (posterior mean; Fig. \ref{fig:dissection_lifetimes_exponential}; Table S\ref{S-tab:dissection_estimated_lifespans}) and across the 131 studies, 95\% of
the individual time series estimates were less than 3 (Fig. \ref{fig:dissection_individualEstimates_allSpecies}). The
estimated greatest number of cycles was for \emph{Anopheles sergentii}
(3.0 cycles; posterior mean) which is adapted to desert conditions (it
is known as the ``oasis vector'' of malaria) and may have evolved
greater longevity. The important African malaria vector \emph{A.
gambiae s.l.} was estimated to be the second longest living (2.4 cycles;
posterior mean). The smallest estimated mean number of gonotrophic
cycles was for \emph{Anopheles} \emph{bellator} (0.6 cycles; posterior
mean) which transmits malaria in Brazil's Atlantic Forest. There were
significant differences in estimated lifetime gonotrophic cycles amongst
species (ANOVA: \emph{F}\textsubscript{24,106} =2.2, \emph{p}
\textless{}0.01; the non-parametric Kruskal Wallace:
\({\chi^{2}}_{24}\), \emph{p}\textless{}0.01).

The estimated lifetime gonotrophic cycles for the different genera were
\emph{Anopheles,} 1.6; \emph{Culex,} 1.2; \emph{Mansonia}, 1.1; and
\emph{Aedes} 0.8 (Fig. \ref{fig:dissection_lifetimes_exponential}; Table S\ref{S-tab:dissection_estimated_lifespans}) and the differences between the genera were
significant (ANOVA: \emph{F}\textsubscript{3,127} =3.4, \emph{p} =0.02;
the non-parametric Kruskal Wallace: \(\chi_{3}^{2} = 21.7\),
\emph{p}\textless{}0.01).

@@

To compare the two methods, we converted numbers of gonotrophic cycle
(physiological age) into lifespan (chronological age) as described in
the SOM using estimates of gonotrophic cycle duration (Fig. \ref{S-fig:dissection_gonotrophicCycleRaw}). Table S\ref{S-tab:dissection_estimated_lifespans_chron} provides posterior summaries of chronological for the species and genera in the dissection dataset. For ten species, we had enough data from both species to make a comparison, and there was a positive correlation (not
statistically significant; Pearson correlation $\rho=0.42$, $n=10$, $p=0.23$)
between the two measures (Fig. \ref{fig:comparison}), and in only one case -- for \emph{A.
darlingi} - there was a significant difference in the
time-series level LBLs (Table S\ref{S-tab:comparison}).

@@

We conducted a power analysis of MRR experiments to determine
whether typical experimental characteristics could detect senescence.
Here we calculated the power of a maximum likelihood estimator of the
`senescence parameter' $\beta$ of the Gompertz survival function (see Table
S\ref{S-tab:mrr_survivalDescription}) for case study populations with three different levels of
senescence (Fig. \ref{S-fig:mrr_mcPowerAnalysis_senescence}A). This analysis indicated that power to detect
senescence strongly depends on study length (Fig. \ref{S-fig:mrr_mcPowerAnalysis_senescence}B) but is
insensitive to release size (Fig. \ref{S-fig:mrr_mcPowerAnalysis_senescence}C). Clements and Patterson (1981)
conducted a meta-analysis of MRR and dissection-based field experiments
and found evidence of an increasing risk of mortality hazard with age
that is similar in magnitude to that of the `mild' case considered
above. For this case, detecting senescence with a power of 80\% requires
a study length of at least 18 days. Since the median study duration for
experiments included in our analysis was 10 days (Table S\ref{S-tab:mrr_IndividualData}) this could
partly explain our failure to detect senescence at the species level. A
number of experiments have found evidence of age-dependence in
laboratory populations \citep{styer2007mosquitoes,dawes2009anopheles}.
However, the artificially benign environment of the laboratory means
mosquitoes live considerably longer than in the wild, where they may die
because of exogenous factors, before the effects of physiological
decline have had time to manifest. Field experiments have also found
evidence for age-dependent mortality. Harrington et al. (2008) conducted
a field experiment where mosquitoes reared under laboratory conditions
were marked and released at different ages. Analysis of the resultant
MRR time-series indicated that mosquito mortality increases with age at
release. It is possible, however, that this field experiment suffers
from the same biases as laboratory-based approaches, because the
released mosquitoes were often of ages considerably higher (up to 20
days) than typical estimates of wild mosquito lifespan.

>>>>

Data from MRR experiments in the Guerra et al. (2014) database were
examined and those with fewer than six recaptures and species with only
a single MRR study were excluded for the hierarchical analysis. Of the
232 data sets, 179 involved only females, 35 males, and 18 both sex
releases. For 102 data sets the age of the released mosquitoes was known
(the average age of released mosquitoes was 4.0 days) while in the other
cases it was unknown or unrecorded; in these cases we assumed the
mosquitoes were newly emerged at the time of release and return to this
assumption later. See Table \ref{S-tab:mrr_aggregateData} for a summary of other data characteristics.

@@

The estimated mean LBL for female mosquitoes not pre-fed with blood or
sugar before release for \emph{Culex, Anopheles} and \emph{Aedes} were
2.9, 6.8 and 8.1 days respectively with an overall estimate of 6.0 days
(Fig. \ref{fig:mrr_lifespans}; Table \ref{S-tab:mrr_estimated_lifespans}). Differences between genera were significant (ANOVA on median
LBL controlling for sex and pre-release feeding:
\emph{F}\textsubscript{2,229} = 12.4, \emph{p}\textless{}0.01; Kruskal
Wallace: \({\chi^{2}}_{2} = 30.8,p < 0.01\)). \emph{K}-fold cross
validation suggests that after the effect of genus is accounted for the
incorporation of a species term provides little predictive power (Fig.
S\#; in part explained by the latter model over-fitting the data where
there are few time series per species).

@@

Dissection allows the number of completed gonotrophic cycles to be
counted and from this the mean number of cycles before death was
estimated. Overall, the mean number of cycles completed in a lifetime
was 1.3 (posterior mean; Fig. \ref{fig:dissection_lifetimes_exponential}; Table \ref{S-tab:dissection_estimated_lifespans}) and across the 131 studies, 95\% of
the individual time series estimates were less than 3 (Fig. \ref{fig:dissection_individualEstimates_allSpecies}). The
estimated greatest number of cycles was for \emph{Anopheles sergentii}
(3.0 cycles; posterior mean) which is adapted to desert conditions (it
is known as the ``oasis vector'' of malaria) and may have evolved
greater longevity. The important African malaria vector \emph{A.
gambiae s.l.} was estimated to be the second longest living (2.4 cycles;
posterior mean). The smallest estimated mean number of gonotrophic
cycles was for \emph{Anopheles} \emph{bellator} (0.6 cycles; posterior
mean) which transmits malaria in Brazil's Atlantic Forest. There were
significant differences in estimated lifetime gonotrophic cycles amongst
species (ANOVA: \emph{F}\textsubscript{24,106} =2.2, \emph{p}
\textless{}0.01; the non-parametric Kruskal Wallace:
\({\chi^{2}}_{24}\), \emph{p}\textless{}0.01).

The estimated lifetime gonotrophic cycles for the different genera were
\emph{Anopheles,} 1.6; \emph{Culex,} 1.2; \emph{Mansonia}, 1.1; and
\emph{Aedes} 0.8 (Fig. \ref{fig:dissection_lifetimes_exponential}; Table \ref{S-tab:dissection_estimated_lifespans}) and the differences between the genera were
significant (ANOVA: \emph{F}\textsubscript{3,127} =3.4, \emph{p} =0.02;
the non-parametric Kruskal Wallace: \(\chi_{3}^{2} = 21.7\),
\emph{p}\textless{}0.01).

@@

To compare the two methods, we converted numbers of gonotrophic cycle
(physiological age) into lifespan (chronological age) as described in
the SOM using estimates of gonotrophic cycle duration (Fig. \ref{S-fig:dissection_gonotrophicCycleRaw}). Table \ref{S-tab:dissection_estimated_lifespans_chron} provides posterior summaries of chronological for the species and genera in the dissection dataset. For ten species, we had enough data from both species to make a comparison, and there was a positive correlation (not
statistically significant; Pearson correlation $\rho=0.42$, $n=10$, $p=0.23$)
between the two measures (Fig. \ref{fig:comparison}), and in only one case -- for \emph{A.
darlingi} - there was a significant difference in the
time-series level LBLs (Table \ref{S-tab:comparison}).

@@

We conducted a power analysis of MRR experiments to determine
whether typical experimental characteristics could detect senescence.
Here we calculated the power of a maximum likelihood estimator of the
`senescence parameter' $\beta$ of the Gompertz survival function (see Table
\ref{S-tab:mrr_survivalDescription}) for case study populations with three different levels of
senescence (Fig. \ref{S-fig:mrr_mcPowerAnalysis_senescence}A). This analysis indicated that power to detect
senescence strongly depends on study length (Fig. \ref{S-fig:mrr_mcPowerAnalysis_senescence}B) but is
insensitive to release size (Fig. \ref{S-fig:mrr_mcPowerAnalysis_senescence}C). Clements and Patterson (1981)
conducted a meta-analysis of MRR and dissection-based field experiments
and found evidence of an increasing risk of mortality hazard with age
that is similar in magnitude to that of the `mild' case considered
above. For this case, detecting senescence with a power of 80\% requires
a study length of at least 18 days. Since the median study duration for
experiments included in our analysis was 10 days (Table \ref{S-tab:mrr_IndividualData}) this could
partly explain our failure to detect senescence at the species level. A
number of experiments have found evidence of age-dependence in
laboratory populations \citep{styer2007mosquitoes,dawes2009anopheles}.
However, the artificially benign environment of the laboratory means
mosquitoes live considerably longer than in the wild, where they may die
because of exogenous factors, before the effects of physiological
decline have had time to manifest. Field experiments have also found
evidence for age-dependent mortality. Harrington et al. (2008) conducted
a field experiment where mosquitoes reared under laboratory conditions
were marked and released at different ages. Analysis of the resultant
MRR time-series indicated that mosquito mortality increases with age at
release. It is possible, however, that this field experiment suffers
from the same biases as laboratory-based approaches, because the
released mosquitoes were often of ages considerably higher (up to 20
days) than typical estimates of wild mosquito lifespan.
